---
title: "Vignette: heteroscedasticity model"
output: md_document
---

This toy example shows how to model *heteroscedasticity in the latent field and the noise*.
It shows : 

- how to initialize the model
- how to diagnose convergence
- how to compare the models (with a stationary model) using DIC
- how to do predictions


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Generate synthetic data
```{r}
set.seed(1)
# Generate locations
locs = cbind(5*runif(3000), 5*runif(3000))
#Generate latent fields for the parameters
latent_field_noise =   GpGp::fast_Gp_sim(c(.5, 1, 1, 0), locs = locs)
latent_field_scale =   GpGp::fast_Gp_sim(c(.5, 1, 1, 0), locs = locs)
# set number of observations
n_obs = 2 * nrow(locs)
# observing with duplicates
observation_idx = c(sample(nrow(locs), nrow(locs) , replace =F), sample(nrow(locs), n_obs - nrow(locs) , replace =T))
observed_locs = locs[observation_idx,]
# covariates
X = as.data.frame(cbind(rbinom(n = n_obs, size = 1, prob = .5)))
colnames(X) = c("binomial")
X_noise = X
# regression coeffs
beta = c(100, 5)
beta_noise = c(-.5, 1)
# get logarithm of the parameters
log_noise_variance = as.matrix(cbind(1, X_noise)) %*% beta_noise + latent_field_noise[observation_idx]
log_scale = as.matrix(latent_field_scale[observation_idx])
# Use NNGP to generate data
NNarray = GpGp::find_ordered_nn(locs, 10) # Nearest Neighbor Array
sparse_chol = Bidart::compute_sparse_chol(covfun_name = "exponential_isotropic", range_beta = log(.1), locs = locs, NNarray = NNarray) # stationary covariance
# get latent NNGP field and observed Gaussian field
latent_field = GpGp::fast_Gp_sim_Linv(Linv = sparse_chol[[1]], NNarray = NNarray)
observed_field = as.vector(exp(.5*log_scale) * latent_field[observation_idx] +exp(.5*log_noise_variance)*rnorm(n_obs)+ cbind(1, as.matrix(X))%*%beta)
```

Let's plot the latent field for the NNGP's marginal variance...

```{r}
Bidart::plot_pointillist_painting(locs, latent_field_scale, main = "field for latent process log variance")
```

Let's plot the latent field for the noise variance...
```{r}
Bidart::plot_pointillist_painting(locs, latent_field_noise, main = "field for noise log variance")
```

Let's plot the latent field...
```{r}
Bidart::plot_pointillist_painting(locs, latent_field, main = "latent field")
```

## Nonstationary model 

Initialize
```{r}
mcmc_nngp_list_nonstat = Bidart::mcmc_nngp_initialize_nonstationary (
  observed_locs = observed_locs, #spatial locations
  observed_field = c(observed_field), # Response variable
  X = X, # Covariates for the observed field
  m = 5, #number of Nearest Neighbors
  reordering = c("maxmin"), #Reordering
  covfun = "exponential_isotropic", response_model = "Gaussian", # covariance model and response model
  noise_X = X_noise, # covariates for the noise
  noise_range = .5, # range for latent field of parameters, if NULL no latent field
  scale_X = NULL, # covariates for the scale
  scale_range = .5, # range for latent field of parameters, if NULL no latent field
  log_NNGP_matern_covfun = "matern_isotropic", # covariance function for the hyperpriors
  log_NNGP_matern_smoothness = 1, # covariance function for the hyperpriors
  n_chains = 3,  # number of MCMC chains
  seed = 10
)
```


Run
```{r, fig.show='hide', results='hide'}
for(i in seq(40))
{
  mcmc_nngp_list_nonstat = Bidart::mcmc_nngp_run_nonstationary(mcmc_nngp_list_nonstat, n_iterations_update = 100, n_cycles = 1)
}
```

Check convergence
```{r}
Bidart::diagnostic_plots(mcmc_nngp_list_nonstat)
```

## Stationary model 

Initialize
```{r}
mcmc_nngp_list_stat = Bidart::mcmc_nngp_initialize_nonstationary (
  observed_locs = observed_locs, #spatial locations
  observed_field = c(observed_field), # Response variable
  X = X, # Covariates for the observed field
  m = 10, #number of Nearest Neighbors
  reordering = c("maxmin"), #Reordering
  covfun = "exponential_isotropic", response_model = "Gaussian", # covariance model and response model
  n_chains = 3,  # number of MCMC chains
  seed = 10
)
```

Run
```{r, fig.show='hide', results='hide'}
for(i in seq(40))
{
  mcmc_nngp_list_stat = Bidart::mcmc_nngp_run_nonstationary(mcmc_nngp_list_stat, n_iterations_update = 100, n_cycles = 1)
}
```

Check convergence
```{r}
Bidart::diagnostic_plots(mcmc_nngp_list_stat)
```


## Model comparison

```{r}
print("DIC nonstat")
print(Bidart::DIC(mcmc_nngp_list_nonstat))
print("DIC stat")
print(Bidart::DIC(mcmc_nngp_list_stat))
```
## Prediction


```{r}
predicted_locs = as.matrix(expand.grid(seq(0, 5, .1), seq(0, 5, .1)))
pred = Bidart::predict_latent_field(mcmc_nngp_list = mcmc_nngp_list_nonstat, predicted_locs = predicted_locs)
```
```{r}
par(mfrow = c(1, 1))
Bidart::plot_pointillist_painting(pred$predicted_locs, pred$summaries$field[1,], cex = 1.8, main = "predicted mean")
Bidart::plot_pointillist_painting(pred$predicted_locs, pred$summaries$field[5,], cex = 1.8, main = "standard deviation of the prediction")
Bidart::plot_pointillist_painting(pred$predicted_locs, pred$summaries$scale_field[1,], cex = 1.8, main = "latent scale field")
```


