---
title: "Vignette range locally anisotropic"
output: md_document
---
  
This toy example shows how to model *locally anisotropic range*.
It shows : 

- how to initialize the model
- how to diagnose convergence
- how to compare the models (with a stationary model) using DIC
- how to do predictions


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Generate synthetic data
```{r}
set.seed(1)  
# locations and latent fields
locs = 5 * matrix(runif(20000), ncol = 2)
locs = locs[GpGp::order_maxmin(locs),]
latent_field_range = cbind(
  GpGp::fast_Gp_sim(c(.5, .5, 1, 0), locs = locs),
  GpGp::fast_Gp_sim(c(.5, .5, 1, 0), locs = locs),
  GpGp::fast_Gp_sim(c(.5, .5, 1, 0), locs = locs))
latent_field_range = latent_field_range %*% matrix(c(1/sqrt(2), 1/sqrt(2), 0, 1/sqrt(2), -1/sqrt(2), 0, 0, 0, 1), 3)

range_beta = c(log(.1)/sqrt(2), log(.1)/sqrt(2), 0)
  
# observing observations, with duplicates
n_obs = 20000
observation_idx = c(sample(10000, 10000 , replace =F), sample(10000, n_obs - 10000 , replace =T))
observed_locs = locs[observation_idx,]
  
# computing sparse chol
NNarray = GpGp::find_ordered_nn(locs, 5)
sparse_chol = Bidart::compute_sparse_chol(covfun_name = "nonstationary_exponential_anisotropic", range_beta = range_beta, locs = locs, NNarray = NNarray, range_field =  latent_field_range, range_X = matrix(rep(1, nrow(locs))))
  
# latent field
latent_field = GpGp::fast_Gp_sim_Linv(Linv = sparse_chol[[1]], NNarray = NNarray)
# observed field
observed_field = as.vector(latent_field[observation_idx] + rnorm(n_obs)* exp(.5 * log(.5))) 
  
```

Let's plot the latent field for the NNGP's range...

```{r}
Bidart::plot_ellipses(locs[seq(200),], latent_field_range[seq(200),], shrink = .01, main = "range_ellipses")
```

Let's plot the latent field...
```{r}
Bidart::plot_pointillist_painting(locs, latent_field)
```

## Nonstationary model 

Initialize
```{r}
mcmc_nngp_list_range = Bidart::mcmc_nngp_initialize_nonstationary (
  observed_locs = observed_locs, #spatial locations
  observed_field = c(observed_field), # Response variable
  m = 5, #number of Nearest Neighbors
  reordering = c("maxmin"), #Reordering
  covfun = "nonstationary_exponential_anisotropic", response_model = "Gaussian", # covariance model and response model
  range_range = .5, 
  log_NNGP_matern_covfun = "matern_isotropic", # covariance function for the hyperpriors
  log_NNGP_matern_smoothness = 1, # covariance function for the hyperpriors
  n_chains = 3,  # number of MCMC chains
  seed = 10
)
```


Run
```{r, fig.show='hide', results='hide'}
for(i in seq(80))
{
  mcmc_nngp_list_range = Bidart::mcmc_nngp_run_nonstationary(mcmc_nngp_list_range, n_iterations_update = 100, n_cycles = 1)
}
```

Check convergence. (Normally diagnostics pop automatically or are all generated using diagnostic_plots() and look great in Rstudio, but in order to get decent pictures in the md lower functions are called)
```{r}
# normally one should use diagnostic_plots()  , but in order to get good pictures lower functions are used
name = "range_log_scale"
Bidart::plot_PSRF(PSRF = Bidart::grb_diags_field(record_arrays = lapply(mcmc_nngp_list_range$records, function(x)x[[name]]), iterations = mcmc_nngp_list_range$iterations$thinning), varname = name, individual_varnames = row.names(mcmc_nngp_list_range$states$chain_1$params[[name]]))
```
```{r}
# normally one should use diagnostic_plots()  , but in order to get good pictures lower functions are used
name = "range_beta"
Bidart::plot_PSRF(PSRF = Bidart::grb_diags_field(record_arrays = lapply(mcmc_nngp_list_range$records, function(x)x[[name]]), iterations = mcmc_nngp_list_range$iterations$thinning), varname = name, individual_varnames = row.names(mcmc_nngp_list_range$states$chain_1$params[[name]]))
```
```{r}
# normally one should use diagnostic_plots()  , but in order to get good pictures lower functions are used
name = "scale_beta"
Bidart::plot_PSRF(PSRF = Bidart::grb_diags_field(record_arrays = lapply(mcmc_nngp_list_range$records, function(x)x[[name]]), iterations = mcmc_nngp_list_range$iterations$thinning), varname = name, individual_varnames = row.names(mcmc_nngp_list_range$states$chain_1$params[[name]]))
```
```{r}
# normally one should use diagnostic_plots()  , but in order to get good pictures lower functions are used
name = "noise_beta"
Bidart::plot_PSRF(PSRF = Bidart::grb_diags_field(record_arrays = lapply(mcmc_nngp_list_range$records, function(x)x[[name]]), iterations = mcmc_nngp_list_range$iterations$thinning), varname = name, individual_varnames = row.names(mcmc_nngp_list_range$states$chain_1$params[[name]]))
```
```{r}
# normally one should use diagnostic_plots()  , but in order to get good pictures lower functions are used
name = "beta"
Bidart::plot_PSRF(PSRF = Bidart::grb_diags_field(record_arrays = lapply(mcmc_nngp_list_range$records, function(x)x[[name]]), iterations = mcmc_nngp_list_range$iterations$thinning), varname = name, individual_varnames = row.names(mcmc_nngp_list_range$states$chain_1$params[[name]]))
```

## Stationary model 

Initialize
```{r}
mcmc_nngp_list_stat = Bidart::mcmc_nngp_initialize_nonstationary (
  observed_locs = observed_locs, #spatial locations
  observed_field = c(observed_field), # Response variable
  m = 5, #number of Nearest Neighbors
  reordering = c("maxmin"), #Reordering
  covfun = "exponential_isotropic", response_model = "Gaussian", # covariance model and response model
  n_chains = 3,  # number of MCMC chains
  seed = 10
)
```

Run
```{r, fig.show='hide', results='hide'}
for(i in seq(40))
{
  mcmc_nngp_list_stat = Bidart::mcmc_nngp_run_nonstationary(mcmc_nngp_list_stat, n_iterations_update = 100, n_cycles = 1)
}
```

Convergence is checked using diagnostic_plots()


## Model comparison

```{r}
print(Bidart::DIC(mcmc_nngp_list_range))
print(Bidart::DIC(mcmc_nngp_list_stat))
```

## Prediction


```{r}
predicted_locs = as.matrix(expand.grid(seq(0, 5, .1), seq(0, 5, .1)))
pred = Bidart::predict_latent_field(mcmc_nngp_list = mcmc_nngp_list_range, predicted_locs = predicted_locs)
```
```{r}
par(mfrow = c(1, 1))
Bidart::plot_pointillist_painting(pred$predicted_locs, pred$summaries$field[1,], cex = 1.8, main = "predicted mean")
Bidart::plot_pointillist_painting(pred$predicted_locs, pred$summaries$field[5,], cex = 1.8, main = "standard deviation of the prediction")
Bidart::plot_ellipses(pred$predicted_locs, log_range = pred$summaries$range_field[1,], shrink = .01, main = "range ellipses (a posteriori mean of log-range parameters)")
```
