---
title: "Toy example range locally anisotropic"
output: html_document
---
  
This toy example shows how to model *locally anisotropic range*.
It shows : 

- how to initialize the model
- how to diagnose convergence
- how to compare the models (with a stationary model) using DIC
- how to do predictions

Sorry about the size of images, GitHub won't let upload big htmls.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Generate synthetic data
```{r}
set.seed(1)
# Generate locations
locs = cbind(5*runif(3000), 5*runif(3000))
#Generate latent fields for the parameters
latent_field_range =  
  cbind(GpGp::fast_Gp_sim(c(.5, .5, 1, 0), locs = locs), 
        GpGp::fast_Gp_sim(c(.5, .5, 1, 0), locs = locs), 
        GpGp::fast_Gp_sim(c(.5, .5, 1, 0), locs = locs))
# observing with duplicates
n_obs = 2 * nrow(locs)
observation_idx = c(sample(nrow(locs), nrow(locs) , replace =F), sample(nrow(locs), n_obs - nrow(locs) , replace =T))
observed_locs = locs[observation_idx,]
# covariates
X = as.data.frame(cbind(rbinom(n = n_obs, size = 1, prob = .5)))
colnames(X) = c("binomial")
# regression coeffs
beta = c(100, 5)
beta_range = matrix(c(-1,.0, .0, -1,.0, .0, 0 ,.0, .0), ncol = 3)
# Use NNGP to generate data
NNarray = GpGp::find_ordered_nn(locs, 10) # Nearest Neighbor Array
sparse_chol = Bidart::compute_sparse_chol(covfun_name = "nonstationary_exponential_anisotropic", range_beta = beta_range, locs = locs, NNarray = NNarray, range_field =  latent_field_range, range_X = as.matrix(cbind(1, locs)))
# get latent NNGP field and observed Gaussian field
latent_field = GpGp::fast_Gp_sim_Linv(Linv = sparse_chol[[1]], NNarray = NNarray)
observed_field =  as.vector(latent_field[observation_idx] +rnorm(n_obs)+ cbind(1, as.matrix(X))%*%beta)
```

Let's plot the latent field for the NNGP's range...

```{r, dpi = 25}
Bidart::plot_ellipses(locs[seq(200),], latent_field_range[seq(200),], shrink = .01, main = "range_ellipses")
```

Let's plot the latent field...
```{r, dpi = 25}
Bidart::plot_pointillist_painting(locs, latent_field)
```

## Nonstationary model 

Initialize
```{r}
mcmc_nngp_list_range = Bidart::mcmc_nngp_initialize_nonstationary (
  observed_locs = observed_locs, #spatial locations
  observed_field = c(observed_field), # Response variable
  X = X, # Covariates for the observed field
  m = 10, #number of Nearest Neighbors
  reordering = c("maxmin"), #Reordering
  covfun = "nonstationary_exponential_anisotropic", response_model = "Gaussian", # covariance model and response model
  range_range = .5, 
  log_NNGP_matern_covfun = "matern_isotropic", # covariance function for the hyperpriors
  log_NNGP_matern_smoothness = 1, # covariance function for the hyperpriors
  n_chains = 3,  # number of MCMC chains
  seed = 10
)
```


Run
```{r, fig.show='hide', results='hide'}
for(i in seq(1))
{
  mcmc_nngp_list_range = Bidart::mcmc_nngp_run_nonstationary(mcmc_nngp_list_range, n_iterations_update = 100, n_cycles = 1)
}
```

Check convergence. Note that you have 3 components for range_beta and 6 components for range_log_scale.
```{r, dpi = 50}
Bidart::diagnostic_plots(mcmc_nngp_list_range)
```

## Stationary model 

Initialize
```{r}
mcmc_nngp_list_stat = Bidart::mcmc_nngp_initialize_nonstationary (
  observed_locs = observed_locs, #spatial locations
  observed_field = c(observed_field), # Response variable
  X = X, # Covariates for the observed field
  m = 10, #number of Nearest Neighbors
  reordering = c("maxmin"), #Reordering
  covfun = "exponential_isotropic", response_model = "Gaussian", # covariance model and response model
  n_chains = 3,  # number of MCMC chains
  seed = 10
)
```

Run
```{r, fig.show='hide', results='hide'}
for(i in seq(1))
{
  mcmc_nngp_list_stat = Bidart::mcmc_nngp_run_nonstationary(mcmc_nngp_list_stat, n_iterations_update = 100, n_cycles = 1)
}
```

Convergence is checked using diagnostic_plots()


## Model comparison

```{r}
print(Bidart::DIC(mcmc_nngp_list_range))
print(Bidart::DIC(mcmc_nngp_list_stat))
```

## Prediction


```{r}
predicted_locs = as.matrix(expand.grid(seq(0, 5, .1), seq(0, 5, .1)))
pred = Bidart::predict_latent_field(mcmc_nngp_list = mcmc_nngp_list_range, predicted_locs = predicted_locs)
```
```{r, dpi = 25}
par(mfrow = c(1, 1))
Bidart::plot_pointillist_painting(pred$predicted_locs, pred$summaries$field[1,,], cex = 1.8, main = "predicted mean")
Bidart::plot_pointillist_painting(pred$predicted_locs, pred$summaries$field[5,,], cex = 1.8, main = "standard deviation of the prediction")
Bidart::plot_ellipses(pred$predicted_locs, log_range = pred$summaries$range_field[1,,], shrink = .01, main = "range ellipses (a posteriori mean)")
```
